Higher-Order unification for free!

### 2024-04-16 paper/main.tex
minor

To reason about unification, we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

\noindent
The second snippet illustrate the deduplication of \linketa. The syntax
<<\elpiIn{N ~$\triangleright$~ G ?- P}>> denotes a $\lambda$Prolog sequent, that is a 
goal \elpiIn{P} under a set \elpiIn{G} of hypothetical rules (introduced
by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
operator) \elpiIn{N} heigenvariables. Sequents are presented to
CHR with their higher order unification variables replaced by
``frozen constants'', that is $A_{xyz}$ becomes <<\elpiIn{uvar ~$f_A$~ [x,y,z]}>>
for a fresh constant $f_A$. Frozen constants are ``defrost'' when
they are part of a new goal (see \cite[section 4.3]{TASSI_2019}).

>>>>

To reason about unification, we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.

% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

\noindent
The second snippet illustrates the deduplication of \linketa. The syntax
<<\elpiIn{N ~$\triangleright$~ G ?- P}>> denotes a $\lambda$Prolog sequent, that is a 
goal \elpiIn{P} under a set \elpiIn{G} of hypothetical rules (introduced
by \elpiIn{=>}) and where the program context binds (via the \elpiIn{pi}
operator) \elpiIn{N} heigenvariables. Sequents are presented to
CHR with their higher order unification variables replaced by
``frozen constants'', that is $A_{xyz}$ becomes <<\elpiIn{uvar ~$f_A$~ [x,y,z]}>>
for a fresh constant $f_A$. Frozen constants are ``defrost'' when
they are part of a new goal (see \cite[section 4.3]{TASSI_2019}).
